Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex

Algorithm~\ref{algo:ranked_expand} shows a 
version of {\sc Expand} where we use the largest unexplained association to
determine where to next add an edge. 
To implement this, we need an effect size
metric by which we can rank effects. Depending on the type of variables, 
we can use the effect size measures shown in Section~\ref{sec:ci_tests}. 
The algorithm selects the edge that has the highest absolute residual 
association, resulting
in prioritization of edges that contribute the most to improving the model.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate this imperfect expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G_{\textrm{true}} \\
				Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G_{\textrm{true}} \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}

where $G_{\textrm{true}}$ is the true data generating DAG.

An important point to note here is that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct. Therefore,
an $ \mathrm{Expert} $ with accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} = \alpha + (1 - \alpha) / 3 $.

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES) on linear Gaussian data. For \textsc{ExpertInLoop}, we
use the \textsc{RankedExpand} method with Pearson's correlation coefficient as
the effect size measure (Section~\ref{sec:ci_tests}). For the analysis, we
simulated data from known DAGs and use the algorithms to recover the original
DAG. We start by generating a random DAG on $10$ nodes and use linear models
with random effects to simulate $ 500 $ samples from it. We simulate each
variable $ V_i $ in the DAG as:

\begin{equation}
	\begin{split}
		\bm{\beta} &= \mathrm{Uniform}([-0.6, 0.6]) \\
		V_i &= \bm{\beta} \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manually and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are "whitelists" or "blacklists"
of edges that must or must not be present in the resulting DAG, or specification
of a temporal structure between the variables. 
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty of our method lies in the interactive
back-and-forth between domain knowledge and data, which provably leads to the correct
result in a polynomial amount of steps (if the expert answers all
ancestral queries correctly).

>>>>

Algorithm~\ref{algo:ranked_expand} shows a 
version of {\sc Expand} where we use the largest unexplained association to
determine where to next add an edge. 
To implement this, we need an effect size
metric by which we can rank effects. Depending on the type of variables, 
we can use the effect sizes shown in Section~\ref{sec:ci_tests}. 
The algorithm selects the edge that has the highest absolute residual 
association, resulting
in prioritization of edges that contribute the most to improving the model.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate an imperfect domain expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:
\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G_{\textrm{true}} \\
				Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G_{\textrm{true}} \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}
where $G_{\textrm{true}}$ is the true data generating DAG.

It is important to note that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct. Therefore,
an $ \mathrm{Expert} $ with accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} = \alpha + (1 - \alpha) / 3 $.

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES) on linear Gaussian data. For \textsc{ExpertInLoop}, we
use the \textsc{RankedExpand} method with Pearson's correlation coefficient as
the effect size (Section~\ref{sec:ci_tests}) and significance threshold
$\alpha=0.05$; note that this means that also the d-separation ``oracle''
makes both type~I and type~II errors. For the analysis, we
simulate data from known DAGs and use the algorithms to recover the original
DAG. We start by generating a random DAG on $10$ nodes and use linear models
with random effects to simulate $ 500 $ samples from it. We simulate each
variable $ V_i $ in the DAG as
$$
	V_i = \bm{\beta} \cdot \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
$$
where each coefficient $\beta_i$ of the vector $\bm{\beta} \in \mathbb{R}^{|\bm{\beta} |}$ is randomly drawn as
$$
\beta_i \sim \mathrm{Uniform}([-0.6, 0.6])
$$

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manually and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are "whitelists" or "blacklists"
of edges 
\citep{scutari2010}, or specification
of a temporal structure between the variables \cite{Bang2023}.
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty 
of our method lies in the interactive
back-and-forth between domain knowledge and data, 
which provably leads to the correct
result in a polynomial amount of steps (if the expert answers all
ancestral queries correctly).
